Free surface computations in mantle convection models

### 2015-08-19 free-surface-paper.tex
Make look slightly more like a standard eigenvalue problem
Fix labels
Fix typo

Finally, we discuss the implementation of a free surface in the open source, community based
mantle convection software \texttt{Aspect}.
\end{abstract}

@@

There have been several approaches to treating real free surfaces in geodynamic simulations.
\citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
treated as an extra variable that was integrated in time. In this approximation, the free surface 
is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface 
corresponding to the hydrostatic pressure perturbation predicted by the surface topography.

A large number of studies have approximated free surfaces in the interior of the domain by using the 
`sticky air' approximation. In this approximation there is a low-viscosity, low-density layer in the fluid 
(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
Typically a free-slip boundary condition is used above the sticky air.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in an out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells.

Recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
participating codes in that benchmark used the `sticky air' approximation for the free surface, where a low viscosity,
low density layer of fluid (termed `air', though its viscosity is much higher) is placed above the free surface, decoupling 
it from the boundary. Typically a free-slip boundary condition is used above the stiky air.

\citet{schmeling2008benchmark} found that the specific properties of the sticky air layer 
controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it introduces.
A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
with a weak crust is important for producing one-sided subduction zones.

@@

Finally, we describe the implementation of a free surface in the mantle convection software \texttt{Aspect}.

@@

For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
velocity $\mathbf{u}$ is zero:
\begin{equation}
- \nabla p_0 = \rho_0 \mathbf{g}
\label{eq:hydrostatic_stokes}
\end{equation}
where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
The total pressure and density can then be decomposed in to variations about their 
reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.

@@

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by the hydrostatic pressure
\begin{equation}
\mathbf{T} \approx \rho_0 g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
Define
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix} 
\end{aligned}
\end{equation}
Then multiplying both sides by $C^{-1}$ we get
\begin{equation}
\tau_i (C^{-1}D)\mathbf{y}_i = \mathbf{y}_i
\label{eq:standard_eigenvalue}
\end{equation}

@@

\begin{equation}
\mathbf{f}_{\mathrm{body}} = \int_{\Omega + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
\label{eq:predict}
\end{equation}

The shape prediction can be approximated by integration of the velocity:
\begin{equation}
\Delta \Omega \approx \theta \Delta t \mathbf{u}
\end{equation}
where $\theta$ is a free parameter that corresponds to the magnitude of the 
correction, where zero is no stabilization and one is fully (quasi) implicit.

@@

\begin{equation}
\int_{\Omega + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
\int_{\Omega} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
\label{eq:kmm}
\end{equation}
The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
The surface integral has the form of a velocity-dependent surface stress pushing down on the 
domain, and can be thought of as an artificial viscous damping of the surface.
Since it depends on the velocity, it can no longer go on the right-hand side of the 
equation, and must be added as a stabilization term in the system matrix.  
Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.

Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
stabilization term. On the reference surface we note that $\mathbf{g} = -g \mathbf{n}$, which
allows us to write the stabilization term as
\begin{equation}
-\theta \Delta t \int_{\partial \Omega} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
\end{equation}
The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.

@@

\begin{equation}
\begin{bmatrix}
A + \theta \Delta t M & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\tau^S_i
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:generalized_eigenvalue}
\end{equation}
where $\tau^S_i$ indicate the eigenvalues of the stabilized system.
This system may be rearranged:
\begin{equation}
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\left(\tau^S_i - \theta \Delta t \right)
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:generalized_eigenvalue}
\end{equation}

@@

The lengthening of relaxation times due to the quasi-implicit stabilization has an unequal 
effect on the modes. They are all lengthened by the same amount, which is a much bigger 
fraction of the total relaxation time for the shorter modes than the longer ones.

@@

One downside of the quasi-implicit scheme is that it requires a modification of the system matrix, 
and as we have seen this shifts its spectrum.
An alternative would be to construct a time-integration scheme with better stability properties.  
The reason that forward Euler integration performs so badly with the decay of topography is that 
at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
to the \emph{other} side of equilibrium, overshooting even more.
We would like to construct an explicit scheme that accounts for following properties which we know the system has:
\begin{itemize}
\item In the absence of forcing, topography always decreases.
\item Relaxation of small amplitude topography takes the form of exponential decay.
\item The decay mode with the shortest relaxation time is the least stable.
\end{itemize}

@@

Again we may take one of the normal modes as an example, differencing the nonstandard finite-difference scheme from the analytical solution:

\begin{equation}
\begin{aligned}
a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
                          &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
\end{aligned}
\end{equation}

Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\ifdetail
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
\end{equation}
\fi
\begin{equation}
a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3)
\end{equation}

@@

\section{Implementation in \texttt{Aspect} }
\label{sec:implementation}
We have implemented the ability to run free surface simulations in the mantle convection software \texttt{Aspect} \citep{kronbichler2012high}. 
\texttt{Aspect} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
which may not have been written yet. In particular, it cannot rely on assumptions regarding the shape of the domain.

Furthermore, \texttt{Aspect} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
\citep[e.g][]{fullsack1995arbitrary,donea2004encyclopedia}.

\texttt{Aspect} is a parallelized, distributed memory code with adaptive mesh refinement.
The free surface implementation works with these features.  We store the mesh vertex positions 
in a fully distributed vector. This vector is continually updated and redistributed across 
processes upon mesh adaptation (which is handled by the adaptive octree library \texttt{p4est}).
We also provide adaptive refinement indicators based on being near to the free surface or 
when the free surface slope is steep to allow for accurate interface tracking.

@@

Figure~\ref{fig:tau_sensitivity} shows in more detail the effects of the choice of $\tau^*$. In a narrow region 
close to the true value of $\tau_\mathrm{min}$ the error becomes very small, but in a broader region nearby 
the errors are larger. Most of the time the enhanced accuracy of the scheme when $\tau^*$ is near to an eigenvalue 
is a curiosity, though it does suggest the possibility for tuning the finite difference scheme to very accurately 
track a specific mode in time.

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
\caption{Sensitivity to $\tau^*$. As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces. The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the almost spectral accuracy of the time marching scheme for that case.}
\label{fig:tau_sensitivity}
\end{figure*}

@@

Finally, we have described the implementation of free surface boundary conditions in 
the open source mantle convection software \texttt{Aspect}. Both the quasi-implicit 
scheme and the nonstandard finite difference scheme are available. The implementation is 
sufficiently general to accommodate many different geometies, rheologies, and 
gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.

>>>>

Finally, we discuss the implementation of a free surface in the open source, community based
mantle convection software \texttt{ASPECT}.
\end{abstract}

@@

There have been several approaches to treating real free surfaces in geodynamic simulations.
\citet{zhong1996free} introduced a pseudo-free-surface formulation, where the free surface coordinate was
treated as an extra variable that was integrated in time. In this approximation, the free surface 
is approximated on the undeformed Eulerian grid by applying pressure boundary conditions on the reference surface.
The pressure is determined by a first-order Taylor series approximation of the hydrostatic pressure profile 
predicted by the surface topography.

A large number of studies have approximated free surfaces in the interior of the domain by using the 
`sticky air' approximation. In this approximation there is a low-viscosity, low-density layer in the fluid 
(termed `air', though its viscosity is much higher) above the free surface, effectively decoupling it from the boundary. 
Typically a free-slip boundary condition is used above the sticky air, though an open boundary may be better \citep{hillebrand2014using}.

Finally, one can use a true free surface, where a stress-free boundary condition is applied on the boundary 
of the domain. In this case, there can be flow in and out of the domain, so the boundary must move in time.
A true free surface has mathematical elegance in that the boundary condition of the domain more closely 
matches the boundary conditions which one is trying to model, but it typically requires a deformable 
domain with frequent remeshing to avoid ill-conditioned cells.

Recently, the nature of surface boundary conditions have been shown to be important for controlling the 
dynamics of subduction zone modeling.  In a benchmark study \citet{schmeling2008benchmark} performed extensive testing on the effect
of a free surface on the sinking of a slab. They found that the nature of the free surface had a large effect 
on the dynamics of the slab, affecting both the shape and the timing of sinking. Most of the 
participating codes in that benchmark used the sticky air approximation.
They found that the specific properties of the sticky air layer 
controlled the shape and timing of the slab, and the viscosity averaging scheme for areas of transitional
composition was extremely important, as the subducting slab entrained significant amounts of the sticky air.
\citet{crameri2012comparison} conducted comparisons between sticky air and true free surface models, 
demonstrating a range of parameters for sticky air which can mitigate some of the difficulties that it introduces.
A study by \citet{quinquis2011role} found that free surface boundary conditions have a large effect on 
trench migration in a subduction zone, and \citet{crameri2012free} found that a free surface combined 
with a weak crust is important for producing one-sided subduction zones.

@@

Finally, we describe the implementation of a free surface in the mantle convection software \texttt{ASPECT}.

@@

For the purposes of this analysis it is useful to define a hydrostatic reference state where the 
velocity $\mathbf{u}$ is zero:
\begin{equation}
- \nabla p_0 = \rho_0 \mathbf{g}
\label{eq:hydrostatic_stokes}
\end{equation}
where $p_0$ is the reference hydrostatic pressure and $\rho_0$ is a reference density profile (which may vary with depth).
The total pressure and density can then be decomposed into variations about their 
reference values: $\rho = \rho_0 + \rho^\prime$, $p = p_0 + p^\prime$.

@@

The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
Rather than analyzing finite deformation to the free surface (a nonlinear problem),
we will make the analytically useful approximation of small deformations about the hydrostatic 
reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce a temporary auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
On the reference surface the gravity vector is opposite the direction of the surface normal $\mathbf{g} = -g \mathbf{n}$.
The stress on this reference surface can be approximated by using the first order Taylor series
expansion of the hydrostatic pressure profile:
\begin{equation}
\mathbf{T} \approx \frac{\partial \mathbf{T}}{\partial \mathbf{n} } \cdot \mathbf{n} \zeta = \rho_0 g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
Equation~\eqref{eq:weak_stokes} then becomes

@@

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
Define
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix} 
\end{aligned}
\end{equation}
Then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
\begin{equation}
(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i
\label{eq:standard_eigenvalue}
\end{equation}

@@

\begin{equation}
\mathbf{f}_{\mathrm{body}} = \int_{\Omega^n + \Delta \Omega} \rho \mathbf{w} \cdot \mathbf{g}
\label{eq:predict}
\end{equation}
where the superscript $n$ on $\Omega$ indicates the shape at the $n$th timestep.
The shape prediction can be approximated by integration of the velocity:
\begin{equation}
\Delta \Omega \approx \theta \Delta t \mathbf{u}
\end{equation}
where $\theta$ is a free parameter that corresponds to the magnitude of the 
correction, where zero is no stabilization and one is fully (quasi) implicit.

@@

\begin{equation}
\int_{\Omega^n + \Delta \Omega} \rho  \mathbf{w} \cdot \mathbf{g} \approx
\int_{\Omega^n} \rho  \mathbf{w} \cdot \mathbf{g} + \theta \Delta t \int_{\partial \Omega^n} \rho ( \mathbf{w \cdot g}) (\mathbf{u \cdot n} )
\label{eq:kmm}
\end{equation}
The volume integral is the same as that of the unstabilized problem, but now there is a surface integral.
The surface integral has the form of a velocity-dependent surface stress pushing down on the 
domain, and can be thought of as an artificial viscous damping of the surface.
Since it depends on the velocity, it can no longer go on the right-hand side of the 
equation, and must be added as a stabilization term in the system matrix.  
Empirically it has been found to be successful at damping instabilities \citep{kaus2010stabilization, quinquis2011role, duretz2011discretization}.

Indeed, using the formalism in Section~\ref{sec:eigenvalue} we can see the effect of this 
stabilization term. On the reference surface we note that $\mathbf{g} = -g \mathbf{n}$, which
allows us to write the stabilization term as
\begin{equation}
-\theta \Delta t \int_{\partial \Omega^n} \rho g ( \mathbf{w \cdot n}) (\mathbf{u \cdot n} )
\end{equation}
The integral here is precisely the same as that on the right hand side of Equation~\eqref{eq:weak_eigen}, which was discretized as the matrix $M$.

@@

\begin{equation}
\begin{bmatrix}
A + \theta \Delta t M & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\tau^S_i
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:stabilized_generalized_eigenvalue}
\end{equation}
where $\tau^S_i$ indicate the eigenvalues of the stabilized system.
This system may be rearranged:
\begin{equation}
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
=
\left(\tau^S_i - \theta \Delta t \right)
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix}
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix}
\label{eq:rearranged_stabilized_generalized_eigenvalue}
\end{equation}

@@

The lengthening of relaxation times due to the quasi-implicit stabilization has an unequal 
effect on the modes. They are all lengthened by the same amount, which is a much bigger 
fraction of the total relaxation time for the shorter-time modes than the longer-time ones.
Therefore the shorter-time modes are effectively damped more.
This can be seen as an attractive feature of the scheme, since those are the least stable modes,
though it can result in less accurate time-marching of the longest wavelengths of the system.

@@

One downside of the quasi-implicit scheme is that it requires a modification of the system matrix.
A naive implementation of the scheme results in a slightly 
asymmetric matrix, which can be more difficult to solve, requiring either
changes to the solver/preconditioner, or symmetrization of the stabilization term \citep{kaus2010stabilization}.
An alternative would be to construct a time-integration scheme with better stability properties.  
The reason that forward Euler integration performs so badly with the decay of topography is that 
at larger timesteps it overshoots its equilibrium position. This overshoot causes it to lurch back 
to the \emph{other} side of equilibrium, overshooting even more.
We would like to construct an explicit scheme that accounts for following properties which we know the system has:
\begin{itemize}
\item In the absence of forcing, topography always decreases.
\item Relaxation of small amplitude topography takes the form of exponential decay.
\item The decay mode with the shortest relaxation time is the least stable.
\end{itemize}

@@

Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
scheme and the analytical solution after one time step:

\begin{equation}
\begin{aligned}
a_i(\Delta t) - a_i^{1} &= a_i(0) e^{-\Delta t/\tau_i} - a_i{(0)} \left[ 1 - \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
                        &= a_i{(0)} \left[ e^{-\Delta t/\tau_i} - 1 + \frac{\tau^*}{\tau_i} \left(1-e^{-\Delta t/\tau^*} \right) \right] \\
\end{aligned}
\end{equation}
Note that when $\tau_i = \tau^*$ the nonstandard finite difference scheme is exact.
The exponentials may be expanded to find
\ifdetail
\begin{equation}
a_i(\Delta t) - a_i^{1} = - a_i{(0)} \left[ \left(\frac{\Delta t}{\tau_i}\right)^2 - \frac{\tau^*}{\tau_i} \left(\frac{\Delta t }{\tau^* }\right)^2 \right] 
\end{equation}
\fi
\begin{equation}
a_i(\Delta t) - a_i^{1} = {a_i{(0)} } \left( \frac{\Delta t}{\tau_i} \right)^2 \left( 1 - \frac{\tau_i}{\tau^*} \right) + O(\Delta t^3)
\end{equation}
Summing this error over many timesteps results in a factor of $\Delta t^{-1}$, demonstrating the first-order accuracy~\citep[e.g.]{leveque2007finite}.
The choice of $\tau^*$ controls the size of the coefficient on the truncation error for the scheme.
The error for a given mode becomes considerably smaller when $\tau^*$ is close to its natural relaxation times.

@@

\section{Implementation in \texttt{ASPECT} }
\label{sec:implementation}
We have implemented the ability to run free surface simulations in the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}. 
\texttt{ASPECT} is designed to be highly flexible and modular, with the ability for user-defined rheologies, geometries, and gravity models. 
The implementation of the free surface, therefore, needs to be general enough to work for many combinations of these models, including those 
which may not have been written yet. In particular, it cannot rely on assumptions regarding the shape of the domain.

Furthermore, \texttt{ASPECT} allows for computations in both 2D and 3D, so the implementation must be sufficiently dimension
independent to work in both cases. We implement the free surface within an arbitrary Lagrangian-Eulerian (ALE) formulation
\citep[e.g][]{fullsack1995arbitrary,donea2004encyclopedia}.

\texttt{ASPECT} is a parallelized, distributed memory code with adaptive mesh refinement.
The free surface implementation works with these features.  We store the mesh vertex positions 
in a fully distributed vector. This vector is continually updated and redistributed across 
processes upon mesh adaptation (which is handled by the adaptive octree library \texttt{p4est}).
We also provide adaptive refinement indicators based on being near to the free surface or 
when the free surface slope is steep to allow for accurate interface tracking.

@@

Figure~\ref{fig:tau_sensitivity} shows in more detail the effects of the choice of $\tau^*$. In a narrow region 
close to the true value of $\tau_\mathrm{min}$ the error becomes very small, but in a broader region nearby 
the errors are larger. 
The excellent accuracy when the stabilization timescale $\tau^*$ is close to one of the natural relaxation
timescales allows for tuning of the scheme to track specific long-wavelength modes, such as 
those due to rotational or tidal deformation.

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/tau_sensitivity.pdf}
\caption{Sensitivity of the nonstandard finite difference scheme to $\tau^*$. As the relaxation parameter $\tau^*$ approaches the relaxation time of the benchmark case the error reduces. The sharp cusp at $\tau^* = \tau_\mathrm{min}$ corresponds to the almost spectral accuracy of the time marching scheme for that case.}
\label{fig:tau_sensitivity}
\end{figure*}

@@

Finally, we have described the implementation of free surface boundary conditions in 
the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
scheme and the nonstandard finite difference scheme are available. The implementation is 
sufficiently general to accommodate many different geometies, rheologies, and 
gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.
